Molecular dynamics simulation of synchronization in driven particles

### 2020-11-12 manuscript/manuscript_ajp_2020.tex
edited language, rearranged intro.

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an externally applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and particle environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  model the particle dynamics 
  using overdamped equations of motion suitable for a viscous suspension
  of microscopic particles.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates
  and the propagation of 
  high density kinks.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We also explore 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  We include sample code
  and exercises for students
  that include 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

\section{Introduction} %
%
%Outline
%------------------------------------------------------------------
%synchronization = widest audience possible
Synchronization is a universal phenomena
that can be modeled with numerical studies of confined,
driven particles.
Christiaan Huygens 1665 observation of
in-phase oscillation of pendulum clocks \cite{Pikovsky2003}
[has been adapted into many useful technologies...]
%Huygens explanation
%that the clocks vibrations are coupled through the supporting
%wall has since been rigorously demonstrated by 20th century
%researchers \cite{Ramirez2016}.
Examples of synchronization occur
in many naturally oscillating systems,
from fireflies signaling in time with a with a pulsed LED \cite{},
to simultaneous flapping of birds wings \cite{},
to neuronal firing in cardiac muscle \cite{} and brain tissue \cite{}.

@@

Recently studies of colloidal particles confined
in optical traps have
been used to examine the microscopic dynamics
of mode-locking when subject to external driving force \cite{juniper2015}.

%-----------------------------------------------
%environments
Particles in confined geometries behave differently than free particles.
Stabilized charged particles form patterns
due to the interplay of the confining environment
and particle interactions.
Narrow channels studies are useful to provide insights 
of how particles move through systems 
such as wires and microchannels.
Biological systems such as
neuron axons and capillaries can also be studied
with these models [more detail and references].
Many such systems execute local oscillations
about stable points [elaborate].
%----------------------------------------------------------
%external forces
An applied external force
increases the diversity of behaviors,
and can cause particles to flow in
a variety of non-linear complex behaviors
including
synchronized, aperiod, or chaotic dynamical patterns.
%
The presence of a modulating surface
can modify these patterns in a variety of ways,
changing the onset of dynamical flows,
and the overall flow patterns.

%-------------------------------------------------------------
%experimental - colloids specifically
Colloidal particles trapped in light fields have proven
a particularly useful medium for studying these behaviors.  %cite 
The relatively large size of the colloids
and ease in control has lead to a rich array of 
experimental results. %THIS IS A TERRIBLE SENTENCE.
Such studies are considered model systems
for experimental systems relatively hard to access and visualize,
such as cold atoms or electron gases. %ELABORATE AND CITE

%another confinement technique:
Acoustic tweezers and surface acoustic wave (SAW) \cite{}
%confines single cells and particles with lower energy than light / lasers
%%X. Ding, J. Shi, S.-C. S. Lin, S. Yazdi, B. Kiraly and T. J. Huang, “Tunable patterning of microparticles and cells using standing surface acoustic waves” Lab Chip, 2012, 12, 2491– 2497.
%%%Acousto-optically generated potential energy landscapes: Potential mapping using colloids under flow Michael P. N. Juniper,1,∗ Rut Besseling,2 Dirk G. A. L. Aarts,1 and Roel P. A. Dullens1

@@

%-----------------------------------------------------------------
%transition from synchronized to chaotic
Disordered chaotic dynamics are also possible,
where irregular, unpredictable time evolution of
nonlinear systems and occurs in mechanical oscillators \cite{chaos}.

%-----------------------------------------------------------
%describe types of physical systems - particle types
Numerical simulations of confined, driven particles
can be used to model a variety of physical phenomena.
Particles which interact over long distances
include colloids, magnetic beads, superconducting vortices, dusty plasmas, electron gases. [more detail and references]
Particles which interact over short distances include
bubble arrays/emulsions [more systems and references].

%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of particle velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
Section~\ref{sec:chaos}
and conclude 
in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
as summarized in Section~\ref{sec:problems}.

@@

\section{Molecular Dynamics Model}
\label{sec:MD}
We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles. 
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$.
Particles are subject to
periodic boundary conditions
such that a particle leaving the edges of the system is mapped
back to a position within the simulation 
by the rules $x_i+L=x_i$ and $y_i+L=y_i$.
We summarize unit units of our simulated variables in Table ~\ref{tab:1}.

\begin{table}[h!]
\centering
\caption{Simulation units defined in the code.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
time &  $\eta / E_0$ \\
length &  $a_0$ \\
velocity &  \\
force & $E_0 / a_0$\\
electric potential & V(r_{ij} = q^2 E_0/r_{ij}), $E_0 = Z^*^2/()4\pi \epsilon \epsilon_0 a_0$ \\
charge/dimensionless interaction strength $q$ \\
effective colloidal charge & $Z^*$ \\
solvent dielectic constant & $\epsilon$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

Particles are subject an external time-dependent driving force
$\vec{F}_{D}(t)$
applied parallel to the y-direction.
We model this force as
\begin{equation}
\vec{F}_{D}(t) = [F_{DC} + F_{AC} \sin(\omega t)] \hat{y},
\end{equation}
with modifiable parameters including
a constant component $F_{DC}$,
and a time dependent component with amplitude $F_{AC}$
and frequency $\omega = 2 \pi f$.
The frequency is scaled in time units,
where [finish].
 We demonstrate how a single particle (Sec.~\ref{sec:one})
 and many particles
 move (Sec.~\ref{sec:sync})
 in response to this applied force in a variety of light fields.
 In Sec.~\ref{sec:kink} we set $F_{DC}$
 to zero and track the motion
 of a high density area of a particle chain
 (i.e. kink dynamics).
 %The driving force does add energy into the system, and some of it is lost.

 We create several model light fields,
 creating a landscape of potential minima and maxima
 that modify the local force on a particle as a function of position
 $\vec{F}_{l,i} = \vec{F}_{l}(\vec{r}_i).$
 The landscape potential $V(\vec{r}) $ are static,
 generating a conservative force $\vec{F}_l= -\nabla V_l(\vec{r})$
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}. 
 One example of a periodic light field is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of troughs in the landscape,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

@@

A single particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location.
Since we model the acceleration $\vec{a}$ as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

\section{Results}
\label{sec:results}

\subsection{Single particle system}
% In subsection headings, only the first word is capitalized.
\label{sec:one}

We place a single particle in a periodic landscape with
corrugations along the y-direction 
\begin{equation}
  \label{eq:landscape}
  V_l(y) = V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
where $N_p=20$ are the number of troughs in the landscape,
and $V_{0x}=2$ and $V_{0x}=2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red regions show local maxima and the blue local minima.
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,

We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$.
The particle hops between troughs in the energy landscape
when the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height.
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

@@

  We acknowledge Harvey Gould and Jan Tobochnik, who supported our development of this article.
  Charles and Cynthia Reichhardt advised 
  the project and provided the molecular dynamics code.
  We acknowledge funding from the M.J. Murdock Charitable Trust
  and the Pacific Research Institute for Science and Mathematics (PRISM).

>>>>

  %What problem did you study and why is it important?
  We study synchronization of interacting particles
  confined to a narrow channel 
  driven by an external applied force.  
  %What methods did you use?
  Using 
  numerical simulations
  we control the particle interactions,
  external force and environment
  in order to mimick experimental studies of driven colloidal particles
  confined by light-fields.
  The molecular dynamics simulations
  model the particle dynamics 
  using overdamped equations of motion suitable for a viscous suspension
  of microscopic particles.
  %What were your principal results?
  We observe
  particle synchronization under a variety of conditions,
  including static and dynamic patterns formed on
  periodic and quasiperiod substrates
  and the propagation of 
  high density kinks.
  %What conclusions have you drawn from these results?
  We demonstrate the transition from
  trapped to sliding dynamics in
  quasiperiodic landscapes differs
  from that of a periodic landscape by [TBD],
  and that kinks of high density propagate
  [more/less ?  ] when [what?].
  We also explore 
  the limits of applied force that
  cause the dynamics to transition to chaotic behavior.
  We include sample code
  and exercises for students
  that include 
  opportunties
  to reproduce our results and propose
  new numerical experiments.
  With only a few particles in two-dimensions,
  the simulation runs quickly,
  making this an appropriate model for undergraduates to explore.

@@

\section{Introduction} %
%
%Outline
%------------------------------------------------------------------
%synchronization = widest audience possible
Synchronization is a universal phenomena
that can be modeled with numerical studies of confined,
driven particles.
%Christiaan Huygens 
%[has been adapted into many useful technologies...]
Examples of synchronization occur
in many naturally oscillating systems,
from Huygens 1665 observation of
in-phase oscillation of pendulum clocks \cite{Pikovsky2003},
to fireflies signaling in time with a with a pulsed LED \cite{},
to simultaneous flapping of birds wings \cite{},
to neuronal firing in cardiac muscle \cite{} and brain tissue \cite{}.

@@

Recently studies of colloidal particles confined
in optical traps have
been used to examine the microscopic dynamics
of mode-locking when subject to external driving force \cite{juniper2015}.
[this is important because...]
%experimental - colloids specifically
Colloidal particles trapped in light fields have proven
a particularly useful medium for studying complex dynamical behaviors.  %cite 
The relatively large size of the colloids
and ease in control has lead to a rich array of 
experimental results. %THIS IS A TERRIBLE SENTENCE.
Such studies are considered model systems
for experimental systems relatively hard to access and visualize,
such as cold atoms or electron gases. %ELABORATE AND CITE

%-----------------------------------------------------------
%particle interaction types
Particles which interact over long distances
include colloids, magnetic beads, superconducting vortices, dusty plasmas, electron gases. [more detail and references]
Particles which interact over short distances include
bubble arrays/emulsions [more systems and references].

%-----------------------------------------------
%environments
Particles in confined geometries behave differently than free particles.
Stabilized charged particles form patterns
due to the interplay of the confining environment
and particle interactions.
Narrow channels studies are useful to provide insights 
of how particles move through systems 
such as wires and microchannels.
Biological systems such as
neuron axons and capillaries can also be studied
with these models [more detail and references].
Many such systems execute local oscillations
about stable points [elaborate].
%----------------------------------------------------------
The presence of a modulating surface
can modify these patterns in a variety of ways,
changing the onset of dynamical flows,
and the overall flow patterns.
%

%external forces
It is often useful to study the
dynamics of particles subject to 
an applied external force,
for instance the flow of charges in a conductor,
or cells responding to a chemical gradient.
The external force 
increases the diversity of dynamical behaviors,
and can cause particles to flow in
a variety of non-linear complex patterns
that may be 
synchronized.
Disordered chaotic dynamics are also possible,
where irregular, unpredictable time evolution of
nonlinear systems and occurs in mechanical oscillators \cite{chaos}.

%-------------------------------------------------------------

%another confinement technique:
%Acoustic tweezers and surface acoustic wave (SAW) \cite{}
%confines single cells and particles with lower energy than light / lasers
%%X. Ding, J. Shi, S.-C. S. Lin, S. Yazdi, B. Kiraly and T. J. Huang, “Tunable patterning of microparticles and cells using standing surface acoustic waves” Lab Chip, 2012, 12, 2491– 2497.
%%%Acousto-optically generated potential energy landscapes: Potential mapping using colloids under flow Michael P. N. Juniper,1,∗ Rut Besseling,2 Dirk G. A. L. Aarts,1 and Roel P. A. Dullens1

@@

%-----------------------------------------------------------------
%transition from synchronized to chaotic

%describe types of physical systems - particle types
In this work,
we perform 
numerical simulations of confined, driven particles
to model a variety of physical phenomena.
%----------------------------------------------------------------------------
%paper outline
In the following paper we describe
our molecular dynamics model in Section~\ref{sec:MD}.
%including various  and confining environments.
We include code to simulate
and visualize the dynamics in this section
and supplementary material.
In Section~\ref{sec:results} we summarize
our results,
including synchronized motion of a single confined particle
driven across a periodic landscape in 
Section~\ref{sec:one} and
multiple interacting particles
in Section~\ref{sec:sync},
including stationary propagation of high density kinks
in Sec.~\ref{sec:kink}.
%we
%demonstrate how uniform environments and applied forces
%and create synchronized flow patterns.
We present these results using standard tools of non-linear oscillators
such phase diagrams of particle velocity versus position.
In Section~\ref{sec:quasiperiod},
we show how an aperiodic landscape modifies the particle dynamics.
Finally we explore the transition to chaotic dynamics in 
Section~\ref{sec:chaos}
and conclude 
in Section~\ref{sec:conclusion}. 
In each section
we suggest exercises for interested students,
as summarized in Section~\ref{sec:problems}.

@@

\section{Molecular Dynamics Model}
\label{sec:MD}
We use a classical two-dimensional model for 
studying the dynamics of $N$ interacting particles. 
Particles are confined in a two-dimensional (2D) 
simulation of area $A = L \times L$ where $L=36.5 a_0$
where $a_0$ is a dimensionless unit of length.
An individual particle $i$ has
position $\vec{r}_i = x_i \hat{x} + y_i \hat{y}$.
Particles are subject to
periodic boundary conditions
such that a particle leaving the edges of the system is mapped
back to a position within the simulation 
by $x_i+L=x_i$ and $y_i+L=y_i$.
We summarize the units of the simulated variables in Table ~\ref{tab:1}.

\begin{table}[h!]
\centering
\caption{Simulation units defined in the code.}
\begin{ruledtabular}
\begin{tabular}{c c p{5cm}}
Quantity & Unit \\
\hline
length &  $a_0$ \\
viscosity/damping constant & $\eta$ \\
dimensionless interaction strength & $q$ \\
effective colloidal charge & $Z^*$ \\
solvent dielectic constant & $\epsilon$ \\
time &  $\eta / E_0$ \\
velocity &  $\eta E_0 / a_0$ \\
force & $E_0 / a_0$\\
electric potential & $V(r_{ij}) = q^2 E_0/r_{ij}$ \\
energy & $E_0 = {Z^*}^2/4\pi \epsilon \epsilon_0 a_0$ \\
\end{tabular}
\end{ruledtabular}
\label{tab:1}
\end{table}

Particles are subject an external time-dependent driving force
$\vec{F}_{D}(t)$
applied parallel to the y-direction.
We model this force as
\begin{equation}
\vec{F}_{D}(t) = [F_{DC} + F_{AC} \sin(\omega t)] \hat{y},
\end{equation}
with modifiable parameters including
a constant component $F_{DC}$,
and a time dependent component with amplitude $F_{AC}$
and frequency $\omega = 2 \pi f$.
%The frequency is scaled in time units,
%where [finish].

 We model several enviroments
 %creating a landscape of potential minima and maxima
 %that modify the force on a particle as a function of position
 $\vec{F}_{l}(\vec{r}) = -\nabla V_l(\vec{r}).$
 The landscape potential $V(\vec{r}) $ are static 
 with fixed minima and maxima
 that are periodic or quasi-periodic,
 as described in Sec.~\ref{sec:results}. 
 One example of a periodic landscape is
 %that used in Sec.~\ref{sec:one}
 where 
 \begin{equation}
   V_l(y) = V_0 \cos{(N_p \pi y / L)}
 \end{equation}
 where $N_p$ are the number of periods,
 and $V_0$ is an adjustable parameter
 to set the depth of the troughs.

@@

A single particle has the equation of motion 
\begin{equation}
\eta \vec{v}_i = \vec{F}_{l,i} + \sum_{i \neq j}^{N} \vec{F}_{ij} + \vec{F}_{D}(t).
\end{equation}
where $\eta = 1$.
The equation of motion provides a direct calculation of the velocity
of an individual particle from its location in the simulation
with respect to other particles $j$.
Since we model the acceleration as zero,
the Verlet method simplifies to 
the Euler method,
which is used to calculate the position at subsequent time steps.
%Most undergraduate students learn the Euler method in Computational Physics and Differential Equations, but don't encounter the Verlet method.
\begin{equation}
\vec{r}_i(t+\Delta t) = \vec{v}_i(t) \Delta t + \vec{r}_i(t).
\end{equation}

\section{Results}
\label{sec:results}
 We demonstrate how a single particle (Sec.~\ref{sec:one})
 and many particles
 move (Sec.~\ref{sec:sync})
 in response to this applied force in a variety of environments.
 In Sec.~\ref{sec:kink} we set $F_{DC}$
 to zero and track the motion
 of a high density area of a particle chain
 (i.e. kink dynamics).
 %The driving force does add energy into the system, and some of it is lost.
%
\subsection{Single particle system}
% In subsection headings, only the first word is capitalized.
\label{sec:one}

We place a single particle in a periodic landscape with
corrugations along the y-direction 
\begin{equation}
  \label{eq:landscape}
  V_l(y) = V_{0y} \cos{(N_p \pi y / L)} 
\end{equation}
where $N_p=20$ are the number of troughs in the landscape,
and $V_{0y}=0.2$.
This is illustrated in Fig.~\ref{fig:1}(a) where
the red (blue) regions show local maxima (minima).
The code for generating
a two dimensional colored plot
of the potential landscape
is calculated by evaluating
the analytic function in Eq.~\ref{eq:landscape}
for a grid of values $(x_n,y_n)$.
%is included in the supplementary material,

%single DC case with figure

%increased DC
We apply an external applied force, 
with 
a constant $F_{AC}$ with frequency $\omega = 2\pi f$
then slowly increase $F_{DC}$
at a rate of [].
The particle hops between troughs in the energy landscape
when the superposition of $F_{AC}$ and $F_{DC}$ is large enough
to overcome the barrier height.
By modifying $F_{DC}$ we 
achieve a variety of oscillation modes.
A mode is a periodic pattern of hops
with a constant average particle velocity, $\bar{v}_{y}$
over a range of driving forces $F_{DC}$.
We illustrate mode-locking in 
the velocity-force plot in Fig.~\ref{fig:1}(b).
Here $\bar{v}_{y}$ is increasing in non-uniform steps,
with a quantized height of
$\bar{v}_{y} = n \lambda f$,
where $n$ is an integer,
$\lambda = S_Y/N_p = 36.5/20 = 1.825$ is the spatial period, or wavelength
of the landscape,
and $f = 0.01$ cycles per time unit.

@@

  We acknowledge Harvey Gould and Jan Tobochnik,
  who invited us to write the article and
  supported its development.
  Charles and Cynthia Reichhardt advised 
  the project and provided the original molecular dynamics code
  written in the C programming language.
  We acknowledge funding from the M.J. Murdock Charitable Trust
  and the Pacific Research Institute for Science and Mathematics (PRISM).
